Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-14 paper.tex

We analyse the inferred evolutionary patterns of the
remaining 630 Pango X designated sequences in the following
subsections.

@@

\begin{table}
% \begin{footnotesize}
\begin{tabular}{lllrl}
\toprule
{} &         parent\_lineages &          pango\_parents &   n &
descendant\_lineages \\
lineage &                         &                        &     &
\\
\midrule
XA      &   [B.1.177.18, B.1.1.7] &     [B.1.1.7, B.1.177] &   5 & \{'XA': 5\} \\
XAA     &          [BA.1, BA.2.9] &         [BA.1*, BA.2*] &   8 &  \{'XAB':
39, 'XAG': 17, 'XAA': 8, 'XQ': 2, 'XU'... \\
XAC     &     [BA.2.3, BA.1.17.2] &  [BA.2*, BA.1*, BA.2*] &  27 &
\{'XAC': 27\} \\
XAE     &            [BA.2, BA.1] &         [BA.2*, BA.1*] &   9 &
\{'XAE': 9\} \\
XAG     &          [BA.1, BA.2.9] &         [BA.1*, BA.2*] &  17 &  \{'XAB':
39, 'XAG': 17, 'XAA': 8, 'XQ': 2, 'XU'... \\
XF      &            [AY.4, BA.1] &    [B.1.617.2*, BA.1*] &   2 &
\{'XF': 2\} \\
XG      &         [BA.1.17, BA.2] &         [BA.1*, BA.2*] &  32 &
\{'XG': 32, 'XAB': 1\} \\
XH      &       [BA.1.20, BA.2.9] &         [BA.1*, BA.2*] &  11 &  \{'XAF':
34, 'XH': 11, 'B.1.1.529': 3, 'XE': 3,... \\
XL      &       [BA.1.17.2, BA.2] &         [BA.1*, BA.2*] &  10 &
\{'XL': 10, 'XAB': 1, 'XU': 1, 'XQ': 1\} \\
XR      &          [BA.1.1, BA.2] &       [BA.1.1*, BA.2*] &   8 &
\{'XQ': 9, 'XR': 8, 'XAB': 2\} \\
XT      &  [BA.2.23, Unknown (R)] &         [BA.2*, BA.1*] &   1 &  \{'BA.2':
1, 'BA.1': 1, 'BA.1.9': 1, 'BA.2.23':... \\
XW      &       [BA.1.1.15, BA.2] &         [BA.1*, BA.2*] &  11 &
\{'XW': 11, 'XN': 2\} \\
XY      &          [BA.1.1, BA.2] &         [BA.1*, BA.2*] &   6 &
\{'XY': 6, 'XAF': 1\} \\
\bottomrule
\end{tabular}
\caption{\label{tab:pango-single-origin}
Table summarising the status of the Pango X-lineages in the
Long ARG that are inferred to have a single recombinant origin
based on the NextClade Pango lineage assignments. Imputed parent lineages
are shown from the ARG and also to parent lineages as defined by
the official Pango designation alias key.
FIXME: if we like this table then we can reformat it so that lists
at the end are spread over multiple lines rather than using Python
dict notation }
\end{table}

@@

The sequences of 13 Pango recombinant lineages trace back to a single recombination
node, as summarised in Table~\ref{tab:pango-single-origin}. We can
see that XA, XAC, XAE, and XF form monophyletic clades, in which all of the
samples given that Pango designation descendand from a single recombination
node.

The details are given in Figure X, where we [give the details of
the numbers of samples, etc for each of these 14 lineages] and
show the lineage subgraph for X?, which makes this simple structure
clear. [Explan the lineage subgraph format a bit]

We exclude XAK it only has a single sample in this dataset, and
that sample traces to an older recombination node that
that is likely to be spurious since it has a very large number of descendants.

[TODO need to be clear if we're looking at single monophyletic origins
here, or whether we include complicated things like XAG etc or not.
Probably not?]

% \item 14 out of 33 of the Pango X lineage inferred to have a single
% recombinant origin. (XA, XAA, XAC, XAE, XAG, XAK, XF, XG, XH, XL,
%  XR, XT, XW, XY)
% \item Total of 148 samples (max 32, average of 10 per group)
% \item XAA and XAG share the same origin, all others are distinct
%     \begin{itemize}
%         \item Ours: BA.1 \& BA.2.9 (6472 -- 8393)
%         \item XAA: BA.1* \& BA.2* (8393 -- 9343)
%         \item XAG: BA.2* \& BA.1* (6516 -- 8392)
%         \item At least one XAB lineage (50 samples, with 10 origins)
%         also descends from this recombinant---unclear what's happening
%     \end{itemize}
% \end{itemize}
% \end{block}
% \end{frame}

>>>>

The remaining remaining 630 Pango X designated sequences trace
back to 50 different recombination nodes. We examine the
inferred inheritance structure in the following subsections.

@@

\begin{table}
% \begin{footnotesize}
\begin{tabular}{lllrl}
\toprule
{} &              parent\_lineages &          pango\_parents &   n &
descendant\_lineages \\
lineage &                              &                        &     &
\\
\midrule
XA      &        [B.1.177.18, B.1.1.7] &     [B.1.1.7, B.1.177] &   5 &
\{'XA': 5\} \\
XAA     &               [BA.1, BA.2.9] &         [BA.1*, BA.2*] &   8 &
\{'XAB': 39, 'XAG': 17, 'XAA': 8, 'XQ': 2, 'XU'... \\
XAC     &          [BA.2.3, BA.1.17.2] &  [BA.2*, BA.1*, BA.2*] &  27 &
\{'XAC': 27\} \\
XAE     &                 [BA.2, BA.1] &         [BA.2*, BA.1*] &   9 &
\{'XAE': 9\} \\
XAG     &               [BA.1, BA.2.9] &         [BA.1*, BA.2*] &  17 &
\{'XAB': 39, 'XAG': 17, 'XAA': 8, 'XQ': 2, 'XU'... \\
XF      &                 [AY.4, BA.1] &    [B.1.617.2*, BA.1*] &   2 &
\{'XF': 2\} \\
XG      &              [BA.1.17, BA.2] &         [BA.1*, BA.2*] &  32 &
\{'XG': 32, 'XAB': 1\} \\
XH      &            [BA.1.20, BA.2.9] &         [BA.1*, BA.2*] &  11 &
\{'XAF': 34, 'XH': 11, 'B.1.1.529': 3, 'XE': 3,... \\
XK      &             [BA.1.1.1, BA.2] &         [BA.1*, BA.2*] &   6 &
\{'XK': 3\} \\
XL      &            [BA.1.17.2, BA.2] &         [BA.1*, BA.2*] &  10 &
\{'XL': 10, 'XAB': 1, 'XU': 1, 'XQ': 1\} \\
XR      &               [BA.1.1, BA.2] &       [BA.1.1*, BA.2*] &   8 &
\{'XQ': 9, 'XR': 8, 'XAB': 2\} \\
XS      &              [AY.36, BA.1.1] &  [B.1.617.2*, BA.1.1*] &   6 &
\{'XS': 4\} \\
XT      &       [BA.2.23, Unknown (R)] &         [BA.2*, BA.1*] &   1 &
\{'BA.2': 1, 'BA.1': 1, 'BA.1.9': 1, 'BA.2.23':... \\
XV      &  [BA.1.1, BA.2, Unknown (R)] &         [BA.1*, BA.2*] &   3 &
\{'BA.2': 8, 'XV': 1\} \\
XW      &            [BA.1.1.15, BA.2] &         [BA.1*, BA.2*] &  11 &
\{'XW': 11, 'XN': 2\} \\
XY      &               [BA.1.1, BA.2] &         [BA.1*, BA.2*] &   6 &
\{'XY': 6, 'XAF': 1\} \\
\bottomrule
\end{tabular}

\caption{\label{tab:pango-single-origin}
Table summarising the status of the Pango X-lineages in the
Long ARG that are inferred to have a single recombinant origin
based on the NextClade Pango lineage assignments. Imputed parent lineages
are shown from the ARG and also to parent lineages as defined by
the official Pango designation alias key.
FIXME: if we like this table then we can reformat it so that lists
at the end are spread over multiple lines rather than using Python
dict notation }
\end{table}

@@

The sequences of 16 Pango recombinant lineages trace back to a single recombination
node, as summarised in Table~\ref{tab:pango-single-origin}.
[FIXME UPDATE THESE]
We can
see that XA, XAC, XAE, and XF form monophyletic clades, in which all of the
samples given that Pango designation descendand from a single recombination
node.

[Discuss remaining]

There is generally good agreement on the imputed parental lineages from
the ARG and the official Pango designations. [Discuss more]
